Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-18 submission/manuscript.Rmd
small edits

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs. Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

In EpiLCs, homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B). We then compared the known functions of germline genes dysregulated only in females (XX only - unique to XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C). Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 - meiotic nuclear division) and egg-specific functions (GO:0007292 - female gamete generation).

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023]. However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for PNC-specific KDM5C target genes. Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E). Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C). However, KDM5C did bind the promoter of 4 out of the 5 genes dysregulated in both the brain and EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_). Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their gene promoters.

@@

Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017]. Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes. However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023]. To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.

@@

Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012]. We first examined how many of our germline-enriched genes had promoter CGIs (TSS &#177; 500 bp) using the UCSC genome browser[@nassarUCSCGenomeBrowser2023]. Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F). CGI-containing germline genes had higher enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G). Germline genes with promoter CGIs were more highly expressed than CGI-free genes across spermatogenesis stages, with highest expression in meiotic spermatocytes (Figure 6H). Contrastingly, CGI-free genes only displayed substantial expression in post-meiotic round spermatids (Figure 6H). Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).

@@

Promoters that showed the most robust loss of CpGme in _Kdm5c_-KO exEpiLCs (lowest q-values) harbored CGIs (Figure 6K). CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test). Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M). Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters. Furthermore, this suggests while KDM5C's cataltyic activity is required for repression of some germline genes, some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me2/3 around the TSS.

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression distinct germline gene classes within somatic lineages. Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. Although sex did not impact whether egg or sperm-specific genes were dysregulated, organismal sex did greatly influence the degree of germline gene dysregulation. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D, which exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007]. However, KDM5D has not been reported to regulate germline gene expression. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated. Thus, ectopic germline gene expression occurs along with overall proper somatic differentiation in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me and CpGme, which are often mutually exclusive[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007]. Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes. Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.

By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation to act as a break against runaway activation of germline-specific programs.

@@

\newpage
# Figures and Tables
- Supplementary table 1: list of all germline genes. 
	- Columns to include: 
		- KDM5C bound vs not
		- Log2fc in EpiLC, brain (separate columns?)
    - CGI vs non

>>>>

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs. Loss of chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

In EpiLCs, homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B). We then compared the known functions of germline genes dysregulated only in females, only in males, or in all samples (Figure 4C). Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 - meiotic nuclear division) and egg-specific functions (GO:0007292 - female gamete generation).

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but not in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023]. However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for PNC-specific KDM5C target genes. Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E). Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C). Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, but genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.

@@

Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017]. Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes. However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.

@@

Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012]. We first examined how many of our germline-enriched genes had promoter CGIs (TSS &#177; 500 bp) using the UCSC genome browser[@nassarUCSCGenomeBrowser2023]. Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F). CGI-containing germline genes had higher enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G). Germline genes with promoter CGIs were more highly expressed than CGI-free genes across spermatogenesis stages, with highest expression in meiotic spermatocytes (Figure 6H). Contrastingly, CGI-free genes only displayed substantial expression in post-meiotic round spermatids (Figure 6H). Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6F).

@@

Promoters that showed the most robust loss of CpGme in _Kdm5c_-KO exEpiLCs (lowest q-values) harbored CGIs (Figure 6K). CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test). Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M). Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters. Furthermore, this suggests while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. While sex did not determine whether egg or sperm-specific genes were dysregulated, it greatly influenced the degree of germline gene dysregulation. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous knockout females also had over double the number of germline DEGs than hemizygous knockout males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D. However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus. Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me2/3 and CpGme, which are generally mutually exclusive[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_, strongly impaired CGI methylation and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline genes employ multiple repressive mechanisms to facilitate CpGme, such as DNMT3A/B recruitment via E2F6 and MGA[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007]. Furthermore, these results indicate the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes. However, further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.

By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that are de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to act as a break against runaway activation of germline-specific programs.

@@

\newpage
# Figures and Tables
